Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-08-30 paper/ms.tex
update ms

\author[0000-0001-7204-6727]{G\'asp\'ar \'A. Bakos}
\affiliation{\princeton}

\author{-----SUGGESTED CONTRIBUTORS, PENDING DISCUSSION, COMMENTS, ANALYSES, ETC-----}
%

\author[0000-0002-3164-9086]{Maximilian N. G\"unther}
\affiliation{\esa}

\author[0000-0003-2058-6662]{George~R.~Ricker}
\affiliation{\mitkavli}

@@

% 250 word max!
\begin{abstract}
%  This work focuses on a subset of low-mass pre-main-sequence stars
%  with rotation periods of 0.2--2 days, optical variability amplitudes
%  of a few percent, and sharp recurring light curve features that last for
%  hundreds of cycles.  The sharp features are probably caused by
%  eclipsing clumps of dust or gas orbiting at the corotation radius.
%  We therefore refer to the objects as ``transient corotating clumps''
%  (TCCs).
%
  Complex rotators (CRs) are low-mass pre-main-sequence stars with light
  curves characterized by periods of 0.2--2 days, amplitudes of a few
  percent, and slowly-evolving sharp features.  The sharp features
  might be caused by eclipsing clumps of dust or gas orbiting at the
  corotation radius.
  To improve our understanding of these
  enigmatic stars, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features may
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of two B stars
  known to have strong multipolar magnetic fields.
\end{abstract}

@@

The dust clump and gas prominence hypotheses are geometrically
indistinguishable; one could imagine dust or gas accumulating in
either of the geometries shown in Figure~\ref{fig:f1}.  The main
distinguishing factor between the two models is the composition of the
occulting material.   ``Dust clumps'' invoke opacity from dust, which
would need to be collisionally charged \citep{2023MNRAS.518.4734S}.
``Gas prominences'' invoke opacity from partially-ionized gas, likely
bound-free transitions in hydrogen or perhaps a molecular opacity.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

In both models, the corotation radius is the location at which matter
concentrates.  This is empirically justified by the observation that
the sharp CR features are superposed over smooth, quasi-sinusoidal
starspot profiles.  The theoretical importance of the corotation
radius has been discussed in studies of rapid magnetic rotators,
including neutron stars \citep[e.g.]{1973ApJ...184..271L}, magnetic
massive stars \citep[e.g.][]{1985Ap&SS.116..285N}, and accreting T
Tauri stars \citep[e.g.][]{2005ApJ...634.1214L}.  In regions where the
magnetic field dominates the flow (i.e.  $B^2/8\pi > \rho v^2 /2$),
matter is dragged along with the field lines.  Assuming a dipolar
magnetic field geometry, then the tendency of both gas and dust to
become trapped at corotation is tied to how of the four relevant
forces -- gravity, Lorentz, inertial Coriolis, and inertial
centrifugal -- the Lorentz and Coriolis only act perpendicular to
field lines, while gravity and the centrifugal force are in balance at
$R_{\rm c}$ \citep[e.g.][Sec.~2]{2005MNRAS.357..251T}.  Another way to
phrase this statement is that there is a local minimum in the
effective potential of the corotating frame; given a flow from the
either the star or from a tenuous accretion disk, this local potential
minimum enables material to build up \citep{2005MNRAS.357..251T}.

While theoretical heritage for understanding rapid magnetic
rotators exists, CRs have remained mysterious because they have been both
hard to discover and hard to characterize.   They are hard to discover
because they are rare: CRs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 50 CRs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CRs are
relatively faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

\begin{figure}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.48\textwidth]{f2completeness.pdf}
		\vspace{-0.5cm}
		\caption{
			{\bf Completeness of the TESS 2-minute cadence data for late-K
      and early M dwarfs near the Sun, from Sectors 1-55.}  
      The orange dotted curve shows the number of stars in successive
      radial shells, each with a width of 10\,pc.  To be counted,
      these stars must meet the listed conditions: they must be red
      ($\bprp$>$1.5$) dwarf stars ($M_{\rm G}$>4) amenable for TESS
      observations ($T$<16).  The blue solid curve shows the fraction
      of such stars with at least one sector of TESS 2-minute cadence
      data acquired between Sectors 1-55.  The latter data comprise
      the selection function for our complex rotator search.
		}
			\vspace{-0.5cm}
		\label{fig:completeness}
	\end{center}
\end{figure}

@@

{\sc Slow growth; rapid death}---LP 12-502 shows at least three
instances in which dips switch off over less than one cycle; we did
not see any such instances of dips switching on.  Dip growth instead
seems to happen more slowly.  For instance, the dip at phase 0-0.1
between cycles 258-290 begins to become detectable during cycle 258,
and growths in depth by about 2\% over the next eight cycles.  The
evolution of this particular dip is most clear in the river plots.
The evolution of the dip group at phases 0.1-0.3 during cycles
1410-1481 is another example of this slow mode of dip growth.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the location of the protoplanetary
disk's truncation radius \citep{2018haex.bookE.142I}.  The disk
truncation radius is traditionally calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep{2016ARA&A..54..135H,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and the innermost planets arriving at $\approx$5-10 stellar
radii before the disk is depleted.

@@

\acknowledgments
LGB is grateful for support from the Heising-Simons 51~Pegasi~b
Fellowship, and for helpful conversations with G.~Laughlin, A.~Mann,
and J.~Spake.  This paper relied on data collected by the TESS
mission, which are publicly available from the Mikulski Archive for
Space Telescopes (MAST).  The specific 120-second cadence observations
can be accessed via
\dataset[DOI]{https://doi.org/10.17909/t9-nmc8-f686}.  Funding for the
TESS mission is provided by NASA’s Science Mission Directorate.  TIC
402980664 in particular was observed at 120-second cadence thanks to
the TESS Guest Investigator programs G022252 (PI: J.~Schlieder;
Sectors 18, 19, 25, 26) and G04168 (PI: R.~Jayaraman; Sector 53).

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
-----
{\bf JNW, MNG, GRR to confirm coauthorship \&/or to help out on interpretation!}
% Pending!
%HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript writing and revision.

>>>>

\author[0000-0001-7204-6727]{G\'asp\'ar \'A. Bakos}
\affiliation{\princeton}
%
%\author{-----SUGGESTED CONTRIBUTORS, PENDING DISCUSSION, COMMENTS, ANALYSES, ETC-----}
%%
%
%\author[0000-0002-3164-9086]{Maximilian N. G\"unther}
%\affiliation{\esa}
%
%\author[0000-0003-2058-6662]{George~R.~Ricker}
%\affiliation{\mitkavli}

@@

% 250 word max!
\begin{abstract}
  We study a set of sub-solar mass pre-main-sequence stars
  with rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp recurring light curve features that last for
  hundreds of cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of variable star as ``complex rotators'' (CRs).
  To improve our understanding of these
  enigmatic stars, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features may
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of B stars
  known to have strong multipolar magnetic fields.
\end{abstract}

@@

The dust clump and gas prominence hypotheses are geometrically
similar; one could imagine dust or gas accumulating in
either of the geometries shown in Figure~\ref{fig:f1}.  The main
distinguishing factor between the two models is the composition of the
occulting material.   ``Dust clumps'' invoke opacity from dust, which
would need to be collisionally charged \citep{2023MNRAS.518.4734S}.
``Gas prominences'' invoke opacity from partially-ionized gas, likely
bound-free transitions in hydrogen or perhaps a molecular opacity.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

In both models, the corotation radius is the location at which matter
concentrates.  The empirical basis for this is that
the sharp CR features are superposed over smooth, quasi-sinusoidal
starspot profiles.  The theoretical importance of the corotation
radius has been discussed in studies of rapid magnetic rotators
\citep[e.g.][]{1973ApJ...184..271L,1985Ap&SS.116..285N,2005ApJ...634.1214L}.
%including neutron stars \citep[e.g.]{1973ApJ...184..271L}, magnetic
%massive stars \citep[e.g.][]{1985Ap&SS.116..285N}, and accreting T
%Tauri stars \citep[e.g.][]{2005ApJ...634.1214L}.  
In regions where the
magnetic field dominates the flow (i.e.~$B^2/8\pi > \rho v^2 /2$),
matter is dragged along with the field lines.  
The tendency of both gas and dust to
become trapped at corotation is then tied to how of the four relevant
forces -- gravity, Lorentz, inertial Coriolis, and inertial
centrifugal -- the Lorentz and Coriolis only act perpendicular to
field lines, while gravity and the centrifugal force are in balance at
$R_{\rm c}$ \citep[e.g.][their Section~2]{2005MNRAS.357..251T}.  Another way to
phrase this statement is that there is a local minimum in the
effective potential of the corotating frame; given a flow from the
either the star or from a tenuous accretion disk, this local potential
minimum enables material to build up \citep{2005MNRAS.357..251T}.

Although theoretical heritage for understanding rapid magnetic
rotators exists, CRs have remained mysterious because they have been both
hard to discover and hard to characterize.   They are hard to discover
because they are rare: CRs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 50 CRs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CRs are
relatively faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, yielding optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

\begin{figure}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.46\textwidth]{f2completeness.pdf}
		\vspace{-0.2cm}
		\caption{
			{\bf Completeness of the TESS 2-minute data for late-K
      and early M dwarfs near the Sun, from Sectors 1-55.}  
      The orange dotted curve shows the number of stars in successive
      radial shells, each with a width of 10\,pc.  To be counted,
      these stars must meet the listed conditions: they must be red dwarf
      stars
      ($\bprp$>$1.5$; $M_{\rm G}$>4) amenable for TESS
      observations ($T$<16).  The blue solid curve shows the fraction
      of such stars with at least one sector of TESS 2-minute cadence
      data acquired between Sectors 1-55.
		}
			\vspace{-0.5cm}
		\label{fig:completeness}
	\end{center}
\end{figure}

@@

{\sc Slow growth; rapid death}---LP 12-502 shows at least three
instances in which dips switch off over less than one cycle; we did
not see any such instances of dips switching on.  Dip growth seems to happen more slowly.  For instance, the dip at phase 0-0.1
between cycles 258-290 begins to become detectable during cycle 258,
and growths in depth by about 2\% over the next eight cycles.  The
evolution of this particular dip is most clear in the river plots.
The evolution of the dip group at phases 0.1-0.3 during cycles
1410-1481 is another example of this slow mode of dip growth.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the location of the protoplanetary
disk's truncation radius \citep{2018haex.bookE.142I}.  The 
truncation radius is traditionally calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep{2016ARA&A..54..135H,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and arriving at $\approx$5-10 stellar
radii before the disk is depleted.

@@

\acknowledgments
LGB is grateful for support from the Heising-Simons 51~Pegasi~b
Fellowship, and for helpful conversations with G.~Laughlin, A.~Mann,
and J.~Spake.  This paper relied on data collected by the TESS
mission, which are publicly available from the Mikulski Archive for
Space Telescopes (MAST).  The specific 120-second cadence observations
can be accessed via DOI\!\!\!~\dataset[10.17909/t9-nmc8-f686]{https://doi.org/10.17909/t9-nmc8-f686}.  Funding for the
TESS mission is provided by NASA’s Science Mission Directorate.  TIC
402980664 in particular was observed at 120-second cadence thanks to
the TESS Guest Investigator programs G022252 (PI: J.~Schlieder;
Sectors 18, 19, 25, 26) and G04168 (PI: R.~Jayaraman; Sector 53).

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
JNW significantly improved the clarity of the manuscript.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
%-----
%{\bf JNW, MNG, GRR to confirm coauthorship \&/or to help out on interpretation!}
% Pending!
%HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript writing and revision.
